The Digital Edition of the Archaic Latin Inscriptions (7th{}-5th century B.C.)

### 2016-02-09 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
corrections files added
fixes to Sarullo

\maketitle
\begin{abstract}
This project consists in the digital edition of the archaic Latin inscriptions (7th – 5th century B.C.) according to the
EpiDoc Guidelines. The edition is the result of an autoptical examination of the epigraphic documents and of the
text-bearing objects, together with the analysis of previous studies. In the particular case of the Forum inscription,
this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § \ref{sec:generalia} and \ref{sec:news};
Giulia Sarullo (GS) is responsible for § \ref{sec:3chart} and Marta Muscariello (MM) for § \ref{sec:4tech}.}
\end{abstract}
\keywords{Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.}

@@

As a matter of fact, face A conforms to the others, showing three vertical dots after \emph{sakros} and also, although less
deep and more worn-out, after \emph{esed}\footnote{The details of significant portions of the inscriptions will be shown on
the website with several photographs. } (Fig.~\ref{fig:1}). This fact was highlighted by \citet{gamurrini_paleografia_1899} and the three dots
appear in the apograph in \citet[col. 1003]{hulsen_neue_1899},  and they could also be guessed in the photograph Anderson 3192
(Archivio Alinari). Nevertheless, the \emph{post} \citet{goidanich_rapporti_1943} \emph{vulgata} and the difficulties in checking in person the
stone, because of its almost unreachable and scarcely illuminated position that led to the publication of studies not
supported by an autoptical check, perpetuated a reading influenced by the uncertainty in distinguishing between natural
cavities or notches due to the nature of the stone and `significant' holes caused by the tool used for writing.

@@

\noindent An epigraphic \emph{corpus} can be digitalized in different ways, according to the specific issues that each project intends to
tackle. Unlike EDR that, as t	he other projects constituting the \emph{Electronic Archive of Greek and Latin Epigraphy}
(EAGLE),\footnote{\url{http://www.eagle-eagle.it/}. The other databases related to EAGLE are the Epigraphic Database Bari
(EDB), \url{http://www.edb.uniba.it/}, the Epigraphische Datenbank Heidelberg (EDH),
\url{http://www.uni-heidelberg.de/institute/sonst/adw/edh/index.html} and Hispania Epigraphica (HE),
\url{http://www.eda-bea.es/}.} is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
Guidelines.\footnote{For further information on EpiDoc and its history, see \citet{Elliot:2006aa}.
The guidelines are available at \url{http://www.stoa.org/epidoc/gl/latest/}.} This is a set of specifications and encoding
tools in XML (\emph{eXtensible Markup Language}) for the scientific edition of ancient texts based on the \emph{Text Encoding Initiative} (TEI), a set of XML specifications designed for the digital publication of texts and manuscripts for
research purposes.\footnote{\url{http://www.tei-c.org/index.xml}. See also \citet{Consortium:2008aa} (About these Guidelines):
``The TEI encoding scheme is of particular usefulness in facilitating the loss-free interchange of data amongst
individuals and research groups using different programs, computer systems, or application software''.} EpiDoc is
becoming more and more a point of reference for digital epigraphic projects\footnote{On the relationship between EAGLE
projects and EpiDoc see \citet{felle_esperienze_2012}.} and it is also the standard chosen for the aggregation of the archives' data in
the recently constituted network, again called EAGLE (\emph{Europeana Network of Ancient Greek and Latin
Epigraphy}),\footnote{\url{http://www.eagle-network.eu}. About the new \emph{Best Practice Network}, co-founded by the European
Commission, see \url{http://www.europeana.eu}.} in which, besides the data of the EAGLE archives, the archaic Latin
inscriptions \emph{corpus} will also converge. As a matter of fact, inscriptions encoded with EpiDoc are not only compatible
with other projects created according to these Guidelines, but they can also be transferred from a system into another
without losing any information; actually, since XML consists in a semantic markup, that is related to the content of
the information and not to its appearance, it is possible to modify the look of the final result by simply operating on
the style sheet,\footnote{In XML, all information related to the formatting of the text are registered on a separate
file called style sheet, see \citet[104, 110-111]{bodard_epidoc:_2009}.} not having to revise the single files. This will facilitate
the integration of the archaic Latin inscriptions into wider digital collections such as EDR and EAGLE-Europeana.
Moreover, since the file thus encoded can also be translated into another markup language, their survival despite any
future technical evolution is guaranteed.\footnote{\citet[37-38]{tissoni_epidoc_2008} and \citet[104-105]{bodard_epidoc:_2009}.}

@@

The major encoding issues concern the rendering of the direction of writing and of reversed and upside-down letters.
Unlike later texts, predominantly left-to-right, the inscriptions of the ILA \emph{corpus} show a certain degree of
fluctuation in the direction of writing. Besides left-to-right, right-to-left and boustrophedic inscriptions, there are
also some particular cases, such as the lamina from Lavinium (CIL I$^2$ 2833) and the Tibur pedestal inscription (CIL I$^2$
2658), that requires a specific treatment. For these texts, it was necessary to adopt the \texttt{{\textless}lb
rend={\textquotedbl}up-to-down{\textquotedbl}{\textgreater}} and \texttt{{\textless}lb
rend={\textquotedbl}down-to-up{\textquotedbl}{\textgreater}} elements in order to render the peculiar directions in
which the text was cut.\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
International Event \emph{Off the beaten track. Epigraphy at the borders} (Bari, September 24th{}-25th, 2015). The discussion
highlighted how this is a matter of great relevance both for the archaic inscriptions and the testimonies from late
antiquity and the necessity to establish a common standard to encode the instances of ``non-standard'' directions of
writing emerged. For this reason, this issue will surely be the object of further consideration in the future.}
Reversed and upside-down letters are usually left unmarked in traditional epigraphic editions and we decided to comply
to this practice. Nevertheless, the \texttt{{\textless}hi{\textgreater}} element has been used to mark up these letters, with
two different values of @rend. For reversed letters, the \texttt{{\textless}hi
rend={\textquotedbl}reversed{\textquotedbl}{\textgreater}} was used, a tag that in the EpiDoc Guidelines is used to
encode the \emph{litterae inversae}, enclosed in double round parentheses, such as in ((C)) for mulier.\footnote{See \citet[1722]{panciera_i_2006}.} For upside-down letters, a new value was provided, \texttt{{\textless}hi
rend={\textquotedbl}upside-down{\textquotedbl}{\textgreater}}, since none of the allowed values of \texttt{@rend} for the
\texttt{{\textless}hi{\textgreater}} element is suitable for this issue.

>>>>

\maketitle
\begin{abstract}
This project consists in the digital edition of the archaic Latin inscriptions (7th – 5th century B.C.) according to the
EpiDoc Guidelines. The edition is the result of an autoptical examination of the epigraphic documents and of the
text-bearing objects, together with the analysis of previous studies. In the particular case of the Forum inscription,
this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § \ref{sec:generalia} and \ref{sec:news};
Giulia Sarullo (GS) is responsible for § \ref{sec:4tech} and Marta Muscariello (MM) for § \ref{sec:3chart}.}
\end{abstract}
\keywords{Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.}

@@

As a matter of fact, face A seems to conform to the others, showing three vertical dots after \emph{sakros}\footnote{The details of significant portions of the inscriptions will be shown on
the website with several photographs. } (Fig.~\ref{fig:1}). This fact was highlighted by \citet{gamurrini_paleografia_1899} and the three dots
appear in the apograph in \citet[col. 1003]{hulsen_neue_1899},  and they could also be guessed in the photograph Anderson 3192
(Archivio Alinari). Nevertheless, the \emph{post} \citet{goidanich_rapporti_1943} \emph{vulgata} and the difficulties in checking in person the
stone, because of its almost unreachable and scarcely illuminated position that led to the publication of studies not
supported by an autoptical check, perpetuated a reading influenced by the uncertainty in distinguishing between natural
cavities or notches due to the nature of the stone and `significant' holes caused by the tool used for writing.

@@

\noindent An epigraphic \emph{corpus} can be digitalized in different ways, according to the specific issues that each project intends to
tackle. Unlike EDR that, as t	he other projects constituting the \emph{Electronic Archive of Greek and Latin Epigraphy}
(EAGLE),\footnote{\url{http://www.eagle-eagle.it/}. The other databases related to EAGLE are the Epigraphic Database Bari
(EDB), \url{http://www.edb.uniba.it/}, the Epigraphische Datenbank Heidelberg (EDH),
\url{http://edh-www.adw.uni-heidelberg.de} and Hispania Epigraphica (HE),
\url{http://www.eda-bea.es/}.} is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
Guidelines.\footnote{For further information on EpiDoc and its history, see \citet{Elliot:2006aa}.
The guidelines are available at \url{http://www.stoa.org/epidoc/gl/latest/}.} This is a set of specifications and encoding
tools in XML (\emph{eXtensible Markup Language}) for the scientific edition of ancient texts based on the \emph{Text Encoding Initiative} (TEI), a set of XML specifications designed for the digital publication of texts and manuscripts for
research purposes.\footnote{\url{http://www.tei-c.org/index.xml}. See also \citet{Consortium:2008aa} (About these Guidelines):
``The TEI encoding scheme is of particular usefulness in facilitating the loss-free interchange of data amongst
individuals and research groups using different programs, computer systems, or application software''.} EpiDoc is
becoming more and more a point of reference for digital epigraphic projects\footnote{On the relationship between EAGLE
projects and EpiDoc see \citet{felle_esperienze_2012}.} and it is also the standard chosen for the aggregation of the archives' data in
the recently constituted network, again called EAGLE (\emph{Europeana Network of Ancient Greek and Latin
Epigraphy}),\footnote{\url{http://www.eagle-network.eu}. About the new \emph{Best Practice Network}, co-founded by the European
Commission, see \url{http://www.europeana.eu}.} in which, besides the data of the EAGLE archives, the archaic Latin
inscriptions \emph{corpus} will also converge. As a matter of fact, inscriptions encoded with EpiDoc are not only compatible
with other projects created according to these Guidelines, but they can also be transferred from a system into another
without losing any information; actually, since XML consists in a semantic markup, that is related to the content of
the information and not to its appearance, it is possible to modify the look of the final result by simply operating on
the style sheet,\footnote{In XML, all information related to the formatting of the text are registered on a separate
file called style sheet, see \citet[104, 110-111]{bodard_epidoc:_2009}.} not having to revise the single files. This will facilitate
the integration of the archaic Latin inscriptions into wider digital collections such as EDR and EAGLE-Europeana.
Moreover, since the file thus encoded can also be translated into another markup language, their survival despite any
future technical evolution is guaranteed.\footnote{\citet[37-38]{tissoni_epidoc_2008} and \citet[104-105]{bodard_epidoc:_2009}.}

@@

The major encoding issues concern the rendering of the direction of writing and of reversed and upside-down letters.
Unlike later texts, predominantly left-to-right, the inscriptions of the ILA \emph{corpus} show a certain degree of
fluctuation in the direction of writing. Besides left-to-right, right-to-left and boustrophedic inscriptions, there are
also some particular cases, such as the lamina from Lavinium (CIL I$^2$ 2833) and the Tibur pedestal inscription (CIL I$^2$
2658), that requires a specific treatment. For these texts, it was necessary to create several new specific elements in order to render the peculiar directions in
which the text was cut.\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
International Event \emph{Off the beaten track. Epigraphy at the borders} (Bari, September 24th{}-25th, 2015). The discussion
highlighted how this is a matter of great relevance both for the archaic inscriptions and the testimonies from late
antiquity and the necessity to establish a common standard to encode the instances of ``non-standard'' directions of
writing emerged. This issue is discussed in \citet{SarulloBari}, where the newly created elements are presented.}
Reversed and upside-down letters are usually left unmarked in traditional epigraphic editions and we decided to comply
to this practice. Nevertheless, the \texttt{{\textless}hi{\textgreater}} element has been used to mark up these letters, with
two different values of @rend. For reversed letters, the \texttt{{\textless}hi
rend={\textquotedbl}reversed{\textquotedbl}{\textgreater}} was used, a tag that in the EpiDoc Guidelines is used to
encode the \emph{litterae inversae}, enclosed in double round parentheses, such as in ((C)) for mulier.\footnote{See \citet[1722]{panciera_i_2006}.} For upside-down letters, a new value was provided, \texttt{{\textless}hi
rend={\textquotedbl}upside-down{\textquotedbl}{\textgreater}}, since none of the allowed values of \texttt{@rend} for the
\texttt{{\textless}hi{\textgreater}} element is suitable for this issue.
